---
title: "Cleaning, manipulating and analyzing of survey data from the Fitbit dataset"
output: html_document
---

## About the data

The FitBit Fitness Tracker Data (CC0: Public Domain, dataset made available through Mobius)is a Kaggle data set that contains personal fitness tracker from thirty fitbit users. Thirty eligible Fitbit users consented to the submission of personal tracker data, including minute-level output for physical activity, heart rate, and sleep monitoring. It includes information about daily activity, steps, and heart rate that can be used to explore users’ habits.

The data was downloaded and processed by Shruti Mukhtyar for Bellabeat Case Study during the week of November 2022 from https://www.kaggle.com/datasets/arashnic/fitbit?select=Fitabase+Data+4.12.16-5.12.16. The dataset contains multiple csv files.

Code required for knitr to work and output figures.
```{r}
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)
knitr::opts_chunk$set(fig.width=9)
```

## Install packages

```{r  load-packages, include=FALSE}
install.packages('tidyverse')
install.packages('ggplot2')
install.packages('lubridate')
library(tidyverse)
library(ggplot2)
library(lubridate)
library(dplyr)
```

## Import csv files

The following code imports multiple csv files from a folder into separate data frames. The data frame names are derived from the csv file names e.g. dailyActivity_merged.csv is imported to the dailyAcitivty dataframe.

```{r import survey data}
folder <- '~/Projects/coursera/Bellabeat Case Study/fitbit_data/'
files <- list.files(path=folder, pattern = '.*csv')
names <- gsub('.csv', '', files)

# Read and store data frames
for(i in 1:length(files)) {
  dfname <- sapply(strsplit(names[i], '_'), getElement, 1)
  assign(paste0(dfname), read.csv(paste0(folder, files[i])))
}
```

## Data validation

### Merge hourly data
The hourly data on Intensities, Calories and Steps is stored in separate files. Create a single dataframe with all the hourly data similar to the daily data. The minute data is not used for this analysis.

```{r}
hourlyActivity <- left_join(hourlyCalories, hourlyIntensities) %>%
  left_join(hourlySteps)
tibble(hourlyActivity)
```

### Add a date time column
```{r}
# Daily activities
dailyActivity$Date <- mdy(dailyActivity$ActivityDate)
hourlyActivity$DateTime <- mdy_hms(hourlyActivity$ActivityHour)
sleepDay$DateTime  <- mdy_hms(sleepDay$SleepDay)
weightLogInfo$DateTime  <- mdy_hms(weightLogInfo$Date)
heartrate$DateTime  <- mdy_hms(heartrate$Time)
```

### Check number of unique records for each dataset.

More people tracked their physical activity compared to their sleep schcedule, weight logs or heartrate. Only 3 users out of 33 are common to all data tables.
```{r}
n_distinct(dailyActivity$Id)
n_distinct(hourlyActivity$Id)
n_distinct(sleepDay$Id)
n_distinct(weightLogInfo$Id)
n_distinct(heartrate$Id)
```
###  Check for Duplicates
```{r}
sum(duplicated(dailyActivity))
sum(duplicated(hourlyActivity))
sum(duplicated(sleepDay))
sum(duplicated(weightLogInfo))
sum(duplicated(heartrate))
```
###  Remove duplicates
```{r}
sleepDay <- distinct(sleepDay)
```

### Verify data integrity

The stat_density function draws a smoother version of the histogram showing the distribution of the values for each daily activity metric. The TotalSteps and TotalDistance histograms show similar distributions, which is expected.
```{r echo = FALSE}
dailyActivity %>%
  pivot_longer(
    cols=TotalSteps:Calories,
    names_to='Metric',
    values_to='Value'
  ) %>%
  ggplot(aes(x = Value)) +  
  stat_density() + 
  facet_wrap(~Metric, scales='free')
```
```{r echo = FALSE}
hourlyActivity %>%
  pivot_longer(
    cols=Calories:StepTotal,
    names_to='Metric',
    values_to='Value'
  ) %>%
  ggplot(aes(y = Value, x = DateTime)) +  
  geom_point() + 
  #scale_x_date(date_labels = "%m-%Y") +
  facet_wrap(~Metric, scales='free')
```
The TotalMinutesAsleep and TotalTimeInBed data show similar clustering. 
```{r echo = FALSE}
sleepDay %>%
  pivot_longer(
    cols=TotalSleepRecords:TotalTimeInBed,
    names_to='Metric',
    values_to='Value'
  ) %>%
  ggplot(aes(y = Value, x = DateTime)) +  
  geom_point() + 
  facet_wrap(~Metric, scales='free')
```

```{r echo = FALSE}
weightLogInfo %>%
  pivot_longer(
    cols=WeightKg:BMI,
    names_to='Metric',
    values_to='Value'
  ) %>%
  ggplot(aes(y = Value, x = DateTime)) +  
  geom_point() + 
  facet_wrap(~Metric, scales='free')
```
The resting heart rate during early hours of morning is lower than during the rest of the day. This also makes sense.
```{r echo = FALSE}
heartrate$Hour <- hour(heartrate$DateTime)
attach(heartrate)
  
heartrate %>%
  group_by(Hour, Value) %>% 
    summarise(N = mean(Value)) %>%
  ggplot(aes(y = Value, x = Hour)) +  
  geom_point()
```

## Data Analysis

### Find most active users
```{r echo = FALSE}
# Find common user ids
Reduce(intersect, list(unique(dailyActivity$Id), unique(hourlyActivity$Id), unique(sleepDay$Id), unique(weightLogInfo$Id), unique(heartrate$Id)))
```
### Using the hourly data explore plot a heat map of average total steps taken in a hour per weekday across all users
```{r echo = FALSE}
# Add new columns with weekdays and week numbers
hourlyActivity$Wday <- wday(hourlyActivity$DateTime, label = TRUE)
hourlyActivity$Hour <- hour(hourlyActivity$DateTime)
attach(hourlyActivity)

#Assign color variables
col1 = '#d8e1cf'
col2 = '#af8dc3'

hourlyActivity  %>%
  select('Id', 'Wday', 'Hour', 'StepTotal') %>%
  group_by(Wday, Hour) %>% 
  summarise(N = mean(StepTotal)) %>% 
  ggplot(aes(Hour, Wday)) + geom_tile(aes(fill = N), colour = "white", na.rm = TRUE) +
    scale_fill_gradient(low = col1, high = col2) +  
    guides(fill=guide_legend(title="Total Steps")) +
    theme_bw() + theme_minimal() + 
    labs(title = "Evenings and Weekends show higher activity levels",
         y = NULL, x = 'Hour') +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```
### Using the daily data plot a heat map of average total steps taken in a week per weekday across all users
```{r echo = FALSE}
# Add new columns with weekdays and week numbers
dailyActivity$Wday <- wday(dailyActivity$Date, label = TRUE)
dailyActivity$Week <- week(dailyActivity$Date)
attach(dailyActivity)

#Assign color variables
col1 = '#d8e1cf'
col2 = '#af8dc3'

dailyActivity  %>%
  select('Id', 'Wday', 'Week', 'TotalSteps') %>%
  group_by(Wday, Week) %>% 
  summarise(N = mean(TotalSteps)) %>% 
  ggplot(aes(Week, Wday)) + geom_tile(aes(fill = N), colour = "white", na.rm = TRUE) +
    scale_fill_gradient(low = col1, high = col2) +  
    guides(fill=guide_legend(title="Total Steps")) +
    theme_bw() + theme_minimal() + 
    labs(title = 'Average Activity Levels do not show much variation week to week',
         y = NULL, x = 'Week') +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

```
### Using the daily data visualize Lightly Active Minutes by weekdays for all unique users.
```{r echo = FALSE}
dailyActivity  %>%
  #filter(Id == '4558609924') %>%
  select('Id', 'Date', 'Wday', 'LightlyActiveMinutes', 'FairlyActiveMinutes') %>%
  group_by(Id, Wday) %>% 
  summarise(N = mean(LightlyActiveMinutes)) %>% 
  ggplot(aes(x=Wday, y=N)) +
  geom_bar(stat="identity", fill="#af8dc3") +
  facet_wrap(~Id) +
  labs(title = 'There is a lot of variation in how active users are during the week',
         y = NULL, x = 'Weekday')
```

```{r echo = FALSE}
dailyActivity  %>%
  #filter(Id == '4558609924' | Id == '5577150313' | Id == '6962181067') %>%
  filter(Week == '17') %>%
  select('Id', 'Wday', 'LightlyActiveMinutes', 'FairlyActiveMinutes', 'VeryActiveMinutes') %>%
  pivot_longer(cols=c('LightlyActiveMinutes', 'FairlyActiveMinutes', 'VeryActiveMinutes'),
                    names_to='Metric',
                    values_to='Minutes') %>%
  group_by(Id, Wday, Metric) %>% 
  summarise(N = sum(Minutes)) %>% 
  ggplot() +
    geom_bar(aes(x=Metric, y=N, fill=Metric), stat="identity") +
    scale_fill_brewer(palette="Accent")+
    labs(title = 'There is a lot of variation in how active FitBit users are during a week',
         y = 'Active Minutes', x = NULL) +
    theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    facet_wrap(~Id)
```